Structure Learning for $\VAR(p)$ Models

### 2019-04-09 paper.tex
Some formal updates and prose.

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A

@@

\section{Theory}
\label{sec:theory}
Consider a zero mean wide sense stationary, vector valued and discrete
time stochastic process $x(t) \in L^n_2(\Omega);\ t \in \Z$, having the matrix
valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, which we assume to be
absolutely summable, and spectra $S(\omega)$:

@@

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\mathcal{H}_{t - 1}$, $\mathcal{H}^{-j}_{t - 1}$
  generated by $\{x_i(t - \tau)\}_{(\tau = 1, i = 1)}^{(\infty, n)}$ and
  $\{x_i(t - \tau)\}_{(\tau = 1, i \ne j)}^{(\infty, n)}$ respectively we have

\begin{equation}
  \linEerr{x_i(t)}{\mathcal{H}_{t - 1}} < \linEerr{x_i(t)}{\mathcal{H}^{-j}_{t - 1}},
\end{equation}

where
$\xi[x \ |\ \mathcal{H}] = \E (x - \linE{x}{\mathcal{H}})^2$ and
$\linE{x}{\mathcal{H}} = \text{proj}_{\mathcal{H}}(x)$ denotes
the (unique) projection onto the Hilbert space $\mathcal{H}$.
\end{definition}

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x(t)$'' is important in that GC relations
can change when components are added to or removed from $x(t)$,
e.g. new GC relations can arise if we remove the observations of a
common cause, and existing GC relations can disappear if we observe a
new mediating series.

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if everything in Gaussian then they are
equivalent \cite{barnett2009granger}.

We continue to review some pertinent facts.

@@

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) z(t - \tau),
\end{equation}

where $c(t)$ is a perfectly predictable sequence\footnote{$c(t)$ is
  perfectly predictable if any sample $c(t_0)$ is enough to determine
  $c(t)$ for every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi, \pi]$}, and
$z(t)$ is an uncorrelated sequence having covariance $I$.  We will assume
that $c(t)$ can be effectively estimated and removed and take
$c(t) = 0$.  Furthermore, we assume that the spectra is uniformly
bounded away from $0$:
$\exists \delta:\ \delta I \preceq S(\omega) \preceq \frac{1}{\delta} I\ \forall \omega$ so that the Wold
representation can be inverted to yield the $VAR(\infty)$ form

@@

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and uniformly bounded spectral density.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\mathcal{H}^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\mathcal{H}_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \not\gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\mathcal{H}_t^{i} \perp \mathcal{H}_{t - 1}^{j}\ |\ \mathcal{H}_{t - 1}^{-j}$}
    \item{\hl{Wold $A(\tau)$ condition}}
    \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982measurement}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\mathcal{H}_{t - 1}^{-j}}) \perp (z -
\linE{z}{\mathcal{H}_{t - 1}^{-j}})\ \forall y \in \mathcal{H}_t^i,\ z \in
\mathcal{H}_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

\begin{proof}
$(1) \implies (2)$: The projection for $x_i(t)$ onto $\mathcal{H}_{t - 1}$ is (by definition) given by a form similar to equation \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

 Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally uncorrelated, it follows that the optimal projection is given by the model coefficients themselves.  This holds similarly for the projection onto $\mathcal{H}_{t - 1}^{-j}$.  Then by $(1)$ we have

@@

$(2) \implies (4)$: In computing $(y - \linE{y}{\mathcal{H}_{t = 1}^{-j}})$ for $y \in \mathcal{H}_t^i$ it is sufficient to consider $y = x_i(t)$ since $\mathcal{H}_{t - 1}^i \subseteq \mathcal{H}_{t - 1}^{-j}$, in which case $(x_i(t) - \linE{x_i(t)}{\mathcal{H}_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows since $v_i(t) \perp \mathcal{H}_{t - 1}$ and $\forall z \in \mathcal{H}_{t - 1}^j\ (z - \linE{z}{\mathcal{H}_{t - 1}^{-j}}) \in \mathcal{H}_{t - 1}$

@@

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality.

@@

We first need to establish some graph theoretic notation and
terminology, collected formally in a definition for the reader's
convenient reference.

\begin{definition}[Causality graph, Parents, etc...]

@@

We refer to the set of \textit{parents} of $x_i$ in $\gcg$ by
$\pa{i}$, that is, $j \in \pa{i} \iff x_j \gc x_i$.

Even though the definitions of Granger-causality naturally suggests
that each node in the graph should have a self-loop (unless the node
is simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  However, we do not
immediately assume that $\gcg$ is a DAG, and so we will allow for $i$
to be it's own \textit{ancestor}.

The set of level $\ell$ \textit{grandparents} of node $i$, denoted
$\gpn{\ell}{i}$, is the set such that $j \in \gpn{\ell}{i}$ if and only if
there is a \textit{directed and cycle-free path} of length $\ell$ in
$\gcg$ from $j$ to $i$ \textit{or} if $j = i$ and there is a cycle of
length $n$ starting at $i$ and ending back at $i$, that is, $i$ can be
it's own grandparent.  Clearly, $\pa{i} = \gpn{1}{i}$.

@@

@@

from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.  We will represent the action
(convolution) of this filter compactly by taking $z^{-1}$ as the
backshift operator when convenient

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n a_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that $a_{ij}(\tau) = 0$ unless
$j \in \anc{i}$.

@@

  \begin{equation*}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation*}

  Now, let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$.  Then from lemma (\ref{lem:adj_matrix}) $S$ is the
  transpose of the adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$
  is non-zero if and only if $j \in \gpn{k}{i}$, and therefore
  $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.
  $\anc{i} = \cup_k\gpn{k}{i}$ we see that $\A(z)_{ij}$ is zero if
  $j \not\in \anc{i}$.

@@

  \begin{align*}
    x_i(t) &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{j \in \anc{i} \setminus \{i\}} \A_{ij}(z) v_j(t)\\
  \end{align*}
\end{proof}

@@

\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \mathcal{H}$ where $\mathcal{H}$ is a Hilbert space having inner product $\langle \cdot, \cdot \rangle$ and let $\{\mathcal{H}_n\}_{n = 0}^\infty$ be a collection of subspaces of $\mathcal{H}$ such that $x \perp \mathcal{H}_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^\infty\mathcal{H}_n} = \linE{x}{\bigvee_{n = 1}^\infty\mathcal{H}_n},
  \end{equation*}

  where $\mathcal{H}_2 \vee \mathcal{H}_2 = \cl \{\alpha + \beta\ |\ \alpha \in \mathcal{H}_1, \beta \in \mathcal{H}_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  We can apply the Gram-schmidt process to the collection of subspaces to obtain a sequence of orthogonal subspaces $\widetilde{\mathcal{H}}_0, \widetilde{\mathcal{H}}_1, \ldots$ with $\widetilde{\mathcal{H}}_0 = \mathcal{H}_0$, $\widetilde{\mathcal{H}}_n \subseteq \mathcal{H}_n$, and $\bigvee_{n = 0}^\infty\mathcal{H}_n = \bigoplus_{n = 0}^\infty \widetilde{\mathcal{H}}_n$, where $\oplus$ indicates the orthogonal sum of subspaces (which is always closed).  Then, we have an explicit representation of the projection

  \begin{align*}
    \linE{x}{\bigvee_{n = 0}^\infty\mathcal{H}_n} &= \sum_{n = 0}^\infty \langle x, x_n \rangle x_n\\
    &\overset{(a)}{=} \sum_{n = 1}^\infty \langle x, x_n \rangle x_n,
  \end{align*}

  where $x_n \in \widetilde{\mathcal{H}}_n \subseteq \mathcal{H}_n$ and $(a)$ follows because of our assumption that $x \perp \mathcal{H}_0$.  This establishes that the projection is infact a member of the smaller subspace $\bigvee_{n = 1}^\infty\mathcal{H}_n$, and since the error can only increase by projecting onto this smaller subspace, we have the conclusion.
\end{proof}

@@

  \begin{align*}
    &\langle \alpha - \linE{\alpha}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
    &{\iff} \langle \alpha + x - \linE{\alpha + x}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(a)}{\iff} \langle \alpha + x, \beta - \linE{\beta}{\mathcal{X}} \rangle = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
    &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
  \end{align*}

  where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\mathcal{H}_2} = \linE{\linE{z}{\mathcal{H}_2}}{\mathcal{H}_1}$ for $\mathcal{H}_2 \subseteq \mathcal{H}_1$.

@@

  \begin{align*}
    \E x_i(t)x_j(t - \tau) &= \E \big(\A_{ii}(z)v_i(t)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{k \in \anc{i}}\E[\big(\A_{ik}(z)v_k(t)\big)v_j(t - \tau)] + \sum_{k \in \anc{j}}\E[v_i(t) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{k \in \anc{i}}\sum_{\ell \in \anc{j}}\E[\big(\A_{ik}(z)v_k(t)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)]\\
  \end{align*}

@@

  However, $k \ne \ell \implies \A_{k\ell}(z) = 0$ (i.e. $\A(z) = I + A(1)z^{-1} + \cdots$) so $\A_{ji}(z)v_i(t - \tau)$ is strictly earlier in time, for $\tau \ge 0$, than $v_i(t)$ and hence $\E[v_i(t)\big(\A_{ji}(z)v_i(t - \tau)\big)] = 0$.  This establishes the fact that $H_t^{(i)} \perp \mathcal{H}_t^{(j)}$ where $H_t^{(i)} = \{\alpha x_i(t)\ |\ \alpha \in \R\}$ is the Hilbert space generated by the current sample of $x_i$.  It is clear that $\mathcal{H}_t^{(i)} = \bigvee_{\tau = 0}^\infty H_{t - \tau}^{(i)}$.

@@

  \begin{align*}
    \linE{x_j(t - s)}{\mathcal{H}_t^{(i)}} &= \linE{x_j(t - s)}{\bigvee_{\tau = 0}^\infty H_{t - \tau}^{(i)}}\\
    &\overset{(a)}{=} \linE{x_j(t - s)}{\bigvee_{\tau = 1}^\infty H_{t - \tau}^{(i)}}\\
    &= \linE{x_j(t - s)}{\mathcal{H}_{t - 1}^{(i)}},
  \end{align*}

@@

  \begin{equation*}
    \mathcal{H}_t^{(i)} \perp \mathcal{H}_{t - 1}^{(j)}\ |\ \mathcal{H}_{t - 1}^{(i)},
  \end{equation*}

@@

\begin{definition}[Simply Causal]
  If the Granger-Causality graph $\gcg$ for $x(t)$ is as a directed
  acyclic graph such that there is at most one non-overlapping
  directed path between any two nodes (excluding self-loops) then we
  will refer to $\gcg$ as being \textit{simply causal}.
\end{definition}

\begin{example}
  A directed tree (more generally directed forest) is a simply causal
  graph.
\end{example}

\begin{lemma}
  If $\gcg$ is simply causal then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \implies j \notin \anc{i}$.
\end{lemma}
\begin{proof}
  Suppose that $\gcg$ is simply causal and that we have both
  $i \pwgc j$ and $j \in \anc{i}$.  We will show that $i \in \anc{j}$,
  which produces a contradiction (since $\gcg$ being a DAG is part of
  the requirements for being simply causal).

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

@@

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will denote the collection vectors of
such random elements as $L_2^n(\Omega)$ or simply $L_2^n$.

We will work with a discrete time and wide-sense stationary process
$x(t)$ taking elements in $L_2$.  We suppose that $x(t)$ has zero
mean, $\E x(t) = 0$, and the matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ is absolutely summable and has a spectra $S(\omega)$:

@@

We will also work frequently with the spaces spanned by the values of
such a process

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  $\H_t^x$ is the space
spanned by the values of $x(t)$ between the ``infinite past'' and
``now'', we will often omit the superscript $x$ which should be clear
from context.  Evidently these spaces are separable and as closed
subspaces of a Hilbert space they are themselves Hilbert.  We will
denote the spaces generated in analagous ways by particular components
of $x$ as e.g. $\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular
components as e.g. $\H_t^{-j}$.

@@

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) w(t - \tau),
\end{equation}

where $c(t)$ is a perfectly predictable sequence\footnote{$c(t)$ is
  perfectly predictable if any sample $c(t_0)$ is enough to determine
  $c(t)$ for every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, and $w(t)$ is an uncorrelated sequence having a diagonal
covariance and $A(0) = I$.  We will assume that $c(t) = 0$, which in
practice is to say that it has been accurately estimated and removed.

Furthermore, we assume that the spectra is uniformly bounded away from
$0$:
$\exists \delta:\ \delta I \preceq S(\omega) \preceq \frac{1}{\delta}
I\ \forall \omega$ so that the Wold representation can be inverted to
yield the $VAR(\infty)$ form

@@

The equations (\ref{eqn:wold}), (\ref{eqn:ar_representation}) can be
represented as $x(t) = \A(z)w(t) = \B(z)x(t) + v(t)$ via the action of
the operators (LTI filters)
$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ and
$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau}$ where the
operator $z^{-1}$ is the backshift operator acting on $\H$.  Since
$||z^{-1}|| = 1$, the inversion formulae

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k
\end{equation}

are valid.
\subsection{Granger Causality}

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\H_{t - 1}$, $\H^{-j}_{t - 1}$

\begin{equation}
  \linEerr{x_i(t)}{\H_{t - 1}} < \linEerr{x_i(t)}{\H^{-j}_{t - 1}},
\end{equation}

where $\xi[x \ |\ \H] = \E (x - \linE{x}{\H})^2$ is the mean squared
estimation error and $\linE{x}{\H} = \text{proj}_{\H}(x)$ denotes the
(unique) projection onto the Hilbert space $\H$.
\end{definition}

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series.

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if everything is Gaussian then they are
equivalent \cite{barnett2009granger}.

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \not\gc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{\hl{Wold $A(\tau)$ condition}}
    \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982measurement}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\H_{t - 1}^{-j}}) \perp (z -
\linE{z}{\H_{t - 1}^{-j}})\ \forall y \in \H_t^i,\ z \in
\H_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

\begin{proof}
$(1) \implies (2)$: The projection for $x_i(t)$ onto $\H_{t - 1}$ is (by definition) given by a form similar to equation \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\H_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

 Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally uncorrelated, it follows that the optimal projection is given by the model coefficients themselves.  This holds similarly for the projection onto $\H_{t - 1}^{-j}$.  Then by $(1)$ we have

@@

$(2) \implies (4)$: In computing $(y - \linE{y}{\H_{t = 1}^{-j}})$ for $y \in \H_t^i$ it is sufficient to consider $y = x_i(t)$ since $\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$, in which case $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows since $v_i(t) \perp \H_{t - 1}$ and $\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t - 1}$

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality.

@@

This notion is of interest for a variety of reasons.  From a purely
theoretical or conceptual standpoint, we will seek to understand when
and how pairwise causality can capture the idea of ``flow of
information'' in the underlying graph.  It will also be useful to
reason about what the conditions under which \textit{unobserved}
components of $x(t)$ may or may not interfere with inference in the
actually observed components.  Finally, we will seek to construct
practical estimation procedures based purely on pairwise causality
tests and understand their properties.

We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

\begin{definition}[Causality graph, Parents]

@@

We refer to the set of \textit{parents} of $x_i$ in $\gcg$ by
$\pa{i}$, that is, $j \in \pa{i} \iff x_j \gc x_i$.
\end{definition}

\begin{definition}[Grandparents, ancestors]
The set of level $\ell$ \textit{grandparents} of node $i$, denoted
$\gpn{\ell}{i}$, is the set such that $j \in \gpn{\ell}{i}$ if and only if
there is a \textit{directed and cycle-free path} of length $\ell$ in
$\gcg$ from $j$ to $i$ \textit{or} if $j = i$ and there is a cycle of
length $\ell$ starting at $i$ and ending back at $i$, that is, $i$ can be
it's own grandparent.  Clearly, $\pa{i} = \gpn{1}{i}$.

@@

Even though the definitions of Granger-causality naturally suggests
that each node in the graph should have a self-loop (unless the node
is simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  However, we do not
immediately assume that $\gcg$ is a DAG, and so \textit{we will allow for $i$
to be it's own ancestor}.

@@

from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.  We will represent the action
(convolution) of this filter compactly by taking $z^{-1}$ as the
backshift operator on $\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$ when
convenient

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n a_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that $\A_{ij}(\tau) = 0$ unless
$j \in \anc{i}$.

@@

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation}

  Now, let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$.  Then from lemma \ref{lem:adj_matrix} $S$ is the transpose
  of the adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$ is
  non-zero if and only if $j \in \gpn{k}{i}$, and therefore
  $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.  The Cayley-Hamilton
  theorem can then be applied to represent the infinite sum of equation
  (\ref{eqn:resolvant_inv}) in terms only of \textit{finite} powers of
  $\B(z)$, so the application of lemma \ref{lem:adj_matrix} in the
  limit is not necessary.  Finally, since $\anc{i} = \cup_{k = 1}^n\gpn{k}{i}$
  we see that $\A(z)_{ij}$ is zero if $j \not\in \anc{i}$.

@@

  \begin{align*}
    x_i(t) &= (I - \B(z))^{-1}\\
    &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{j \in \anc{i} \setminus \{i\}} \A_{ij}(z) v_j(t)\\
  \end{align*}
\end{proof}

@@

\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a Hilbert space having inner product $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^\infty$ be a collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} = \linE{x}{\bigvee_{n = 1}^\infty\H_n},
  \end{equation*}

  where $\H_2 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  We can apply the Gram-schmidt process to the collection of subspaces to obtain a sequence of orthogonal subspaces $\widetilde{\H}_0, \widetilde{\H}_1, \ldots$ with $\widetilde{\H}_0 = \H_0$, $\widetilde{\H}_n \subseteq \H_n$, and $\bigvee_{n = 0}^\infty\H_n = \bigoplus_{n = 0}^\infty \widetilde{\H}_n$, where $\oplus$ indicates the orthogonal sum of subspaces (which is always closed).  Then, we have an explicit representation of the projection

  \begin{align*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} &= \sum_{n = 0}^\infty \inner{x}{x_n} x_n\\
    &\overset{(a)}{=} \sum_{n = 1}^\infty \inner{x}{x_n} x_n,
  \end{align*}

  where $x_n \in \widetilde{\H}_n \subseteq \H_n$ and $(a)$ follows because of our assumption that $x \perp \H_0$.  This establishes that the projection is infact a member of the smaller subspace $\bigvee_{n = 1}^\infty\H_n$, and since the error can only increase by projecting onto this smaller subspace, we have the conclusion.
\end{proof}

@@

  \begin{align*}
    &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
    &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
    &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
  \end{align*}

  where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.

@@

  \begin{align*}
    \E x_i(t)x_j(t - \tau) &= \E \big(\A_{ii}(z)v_i(t)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{k \in \anc{i}}\E[\big(\A_{ik}(z)v_k(t)\big)v_j(t - \tau)] + \sum_{k \in \anc{j}}\E[v_i(t) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{k \in \anc{i} \setminus i}\sum_{\ell \in \anc{j} \setminus j}\E[\big(\A_{ik}(z)v_k(t)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)]\\
  \end{align*}

@@

  However, $k \ne \ell \implies \A_{k\ell}(z) = 0$ (i.e. $\A(z) = I + A(1)z^{-1} + \cdots$) so $\A_{ji}(z)v_i(t - \tau)$ is strictly earlier in time, for $\tau \ge 0$, than $v_i(t)$ and hence $\E[v_i(t)\big(\A_{ji}(z)v_i(t - \tau)\big)] = 0$.  This establishes the fact that $H_t^{(i)} \perp \H_t^{(j)}$ where $H_t^{(i)} = \{\alpha x_i(t)\ |\ \alpha \in \R\}$ is the Hilbert space generated by the current sample of $x_i$.  It is clear that $\H_t^{(i)} = \bigvee_{\tau = 0}^\infty H_{t - \tau}^{(i)}$.

@@

  \begin{align*}
    \linE{x_j(t - s)}{\H_t^{(i)}} &= \linE{x_j(t - s)}{\bigvee_{\tau = 0}^\infty H_{t - \tau}^{(i)}}\\
    &\overset{(a)}{=} \linE{x_j(t - s)}{\bigvee_{\tau = 1}^\infty H_{t - \tau}^{(i)}}\\
    &= \linE{x_j(t - s)}{\H_{t - 1}^{(i)}},
  \end{align*}

@@

  \begin{equation*}
    \H_t^{(i)} \perp \H_{t - 1}^{(j)}\ |\ \H_{t - 1}^{(i)},
  \end{equation*}

@@

% \begin{definition}[Simply Causal]
%   If the Granger-Causality graph $\gcg$ for $x(t)$ is as a directed
%   acyclic graph such that there is at most one non-overlapping
%   directed path between any two nodes (excluding self-loops) then we
%   will refer to $\gcg$ as being \textit{simply causal}.
% \end{definition}

% \begin{example}
%   A directed tree (more generally directed forest) is a simply causal
%   graph.
% \end{example}

\begin{lemma}
  If $\gcg$ is a strongly causal DAG then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \implies j \notin \anc{i}$.
\end{lemma}
\begin{proof}
  Suppose that $\gcg$ is a strongly causal DAG and that we have both
  $i \pwgc j$ and $j \in \anc{i}$.  We will show that $i \in \anc{j}$,
  which produces a contradiction (since there must then be a directed cycle).
